# Chapter 6) Analysis of longitudinal data

```{r}
date()
```


```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lme4)
```

## RATS

### Overview of the data

The data consists of longitudinal measurements of weigth in three groups of rats following different diets. 

```{r}
RATSL <- read.csv("C:/Users/labpaavo/IODS-project/data/RATSL.csv")

str(RATSL)

```

*ID* and *group* need to be changed to be changed to factor again.

```{r}
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

glimpse(RATSL)
```


```{r}
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") + theme(legend.position = "top")
```

We see that there are large differences in the starting weights between the groups, group3 having clearly animals with the lowest weight. 

### summary measures approach  

```{r}
# plotting the data 
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") + theme(legend.position = "top")

# groups differ a lot in their starting weight -> standardizing the data 
RATSLS <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# plot the standardized data 
ggplot(RATSLS, aes(x = Time, y = stdweight, group = ID)) +
  geom_line(aes(linetype = Group)) + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "std Weight") + theme(legend.position = "top")

```

Looking at the standardized data, there doens't seem to be clear changes in weights in any of the groups. 

### Mean response profiles 

```{r}
RATS_mean <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight) ) %>%
  ungroup()

ggplot(RATS_mean, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")
```

Mean response profiles show that weight increases slightly in all of the groups and that the weight differences between groups seem to remain throughout the study. 

### Boxplots

```{r}
# summary data ignoring baseline week 

RATS_sum <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

ggplot(RATS_sum, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight)")
```

Each group shows one outlier. Next I will remove the outliers. 

```{r}
# filter outliers 
RATS_filtered <- filter(RATS_sum, 
                    mean > 250 & Group == '1' | 
                    mean < 500 & Group == '2' | 
                    mean > 500 & Group == '3')

# draw the plot again
ggplot(RATS_filtered, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight)")
```

With the outliers removed, there seems to be clear differences in the mean weight between groups. 

### Comparing groups with t-test

```{r}
# t-test for all the groups
for(i in list(c('1', '2'), c('2', '3'), c('1', '3'))){
    print(t.test(mean ~ Group, 
    data = filter(RATS_filtered, Group == i[1] | Group == i[2]), 
    var.equal = TRUE))
}
```

T-test confirms that all the groups differ significantly from each other in regards to mean weight (p-value <.001).

### Linear model with baseline as covariate

```{r}
# baseline from original data

# I only imported long data but I need wide data to get column WD1
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

RATS_sum2 <- RATS_sum %>% mutate(baseline = RATS$WD1)

# linear model 
fit <- lm(mean ~ baseline + Group, data = RATS_sum2)
anova(fit)

```

Linear regression shows that the baseline is the strongest predictor of weight in the experiment, and not the group. This can be seen in the previous figures as the rate of increase in weight between groups did not seem to differ significantly. 

## BRPS

### Overview of the data

The data consists of longitudinal data on brief psychiatric rating scale (BPRS) results in two treatment groups. 

```{r}
BPRSL <- read.csv("C:/Users/labpaavo/IODS-project/data/BPRSL.csv")

str(BPRSL)
```

*treatment* and *subject* need to be changed to be changed to factor again.

```{r}
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

glimpse(BPRSL)
```

### Linear mixed models to BPRS data

```{r}
# plotting raw data
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

The raw data shows that the bprs score seems to decrease in both treatment groups over time. 

```{r}
# regular linear regression model
BPRS_reg <- lm(bprs ~ week + treatment, BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```

The regular linear regression model shows week as a significant predictor for bprs (p<.001), which can be seen already from the plot of the raw data. On the other hand, treatment group does not seem to predict bprs score (p = 0.661). However, regular linear regression assumes that the repeated measures of *bprs* are indenpendent, which is highly unlikely. 

```{r}
# Random intercept model 

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

Now that linear fit for each individual can have different intercept the model takes into account that repeated measures are not necessarily independent. However, comparing linear model and linear intercept model the estimates and standard errors for week and treatment are very similar. This could suggest that indeed treatment does not predict bprs in this data. 

```{r}
# Random slope model

BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
# print a summary of the model
summary(BPRS_ref1)

# ANOVA
anova(BPRS_ref1, BPRS_ref)

```

The test result shows that random intercept and slope model gives a chi-squared statistic of 7.27 with 2 degrees of freedom (p-value of 0.03). This suggests that random intercept and slope model offers a better fit than the random intercept model.

```{r}
# Random intercept and slope model with week x treatment interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + week * treatment + (week | subject), data = BPRSL, REML = FALSE)
# print a summary of the model
summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)
```

The interaction model gives Chisq of 3.17 with 1 degree of freedom (p value of 0.07). Thus, it seems that interaction does not further improve the previous model without interaction.

### Plotting observed and fitted values

```{r}
# draw the plot of BPRSL with the observed Weight values
ggplot(BPRSL, aes(x = week, y = bprs, color = treatment)) +
  geom_line(aes(linetype = subject)) +
  scale_x_continuous(name = "Week") +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

# Create a vector of the fitted values, BPRS_ref1 offered best fit
Fitted <- fitted(BPRS_ref1)

# draw the plot of RATSL with the Fitted values of weight
library(ggplot2)
ggplot(BPRSL, aes(x = week, y = Fitted, color = treatment)) +
  geom_line(aes(linetype = subject)) +
  scale_x_continuous(name = "Week") +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

```

Based on the plot (observed and fitted values), there does not seem to be differences between the treatment groups. Overall, we see that bprs decreases over time, independent to which treatment group the subject belonged. Based on this data one could not recommend one treatment over the other (when measured with bprs).     


